Discussion Draft

### 2020-04-15 doc/Discussion.rmd
more sentence changes and adding citations

Our results of greater between cell variation in females are supported by empirical results from human and mouse [@lynn2002, @gruhn2013]. While the modifier models parameterize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant.

@@

Regardless of whether these models and hypotheses are true, our results further support the results of females having greater variation in the number of crossovers across cells. This feature of heterochiasmy supports evolutionary theories predicting distinct reproductive strategies between females (diversifying offspring) and males (maintain successful haplotypes) [@trivers2002].

@@

As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a highly conserved trait, which our results support. The two locus modifier and SACE models predict sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks under linkage in males respectively.

The COM model, predicts the sex differences in the recombination landscape via interference is due to a combination of the length of the axis and differing strengths of chromosome movements during the pairing process [@hulten2011_COM, @rubin2020mixing].

@@

Our results from musculus^MSM^ and musculus^PWD^ show that the SC length and chromatin compaction is uncoupled from the direction of heterochiasmy. These results slightly depart from predictions which nominate chromatin organization (axis length) as the primary driver of heterochiasmy [@petkov2007]. While the relationship may hold for more general heterochiasmy patterns across broad scale comparisons, these results indicate that rapid male specific  evolution in gwRR, is through a different trait than chromatin compaction.

@@

Examples of evolution of interference strength are rare and may be complicated by changes in the underlying karyotype structure (Segura et al 2014). The general pastern of empirical measures of interference strength is that it has a negative correlation to the genome wide recombination rate [@ottoPaysuer2019]. This fits the logic of the chromosome axis acting as the 'real estate' for DSBs and aligns with known molecular mechanisms (cite). Two examples of the opposite direction, a positive correlation of interference strength and genome-wide recombination rate, involve descriptions of observations at the single bivalent level; the between lab-raised and wild mice of *Peromyscus leucopus* from [@peterson2019] or large effects on genome wide recombination rates across a short evolutionary differences; the within-sex comparison of two breeds of cattle with different genome wide recombination rates [@ma2015_cattle]. We propose that there is a difference in resolution and power between these two groups of empirical results with the positive and negative relationship between genome wide recombination rates and interference strength.

We propose that the **spindle based selection model** can explain the positive correlation and evolution of stronger interference strength via selection on the amount of sister cohesion connecting homologs (figure X, table X). The evolution of genome wide recombination rates in our study is driven by a transition of recombination landscapes from single crossovers to two crossovers. All else being equal, interference strength would be expected to be equal or even weaker. Our results show that the two crossovers are spaced further apart when SC length is controlled for and chromosome size effects are minimized, in the strain with more two crossover chromosomes. Assuming that chromatin compaction between pachtyene and metaphase is uniform along chromosomes, an outcome of this further spacing of two foci results in an increased area of sister cohesion connecting homologs (Figure X). This model gives rise to the possibility that the chromosome structure at metaphase is the target of selection  which is achieved by changing the average recombination landscape across bivalents within the cell. This interpretation has potential to fit with hypotheses mentioned above for explaining other results regarding the evolution of the recombination landscape.

The **COM** model predicts that interference and the recombination landscape arises from known oscillatory movements during prophase, however it lacks evolutionary based predictions. While the **SACE** modifier model does not explicitly model evolution of interference strength we note that a logical outcome of the main prediction of maintaining larger chromosome blocks in males, would be a landscape with stronger interference strength.

We have focused on the model involving direct selection on the recombination pathway, that is the impact on the meiotic process on an individual, but we make note on models centering on indirect selection or fitness outcomes of offspring which could also explain our results. Models from Goldstein et al, reviewed in [@ottoPaysuer2019] and Veller 2017, predict that interference strength evolves whenever increased recombination rates evolve. The simulations under this model indicate that the space across multiple loci or between multiple crossovers increases in a positive manner with the genome wide recombination rate. A distinguishing feature of their model is that the number of crossovers per chromosome is kept constant. The range of crossovers per chromosome is quite small, 1-3, for the majority of chromosomes [@ottoPaysuer2019, @stapley_variation_2017]. Constraining the number of crossovers per chromosome in models may fit empirical data better than those where recombination rate across an abstracted genetic space is unconstrained.

>>>>

<While the modifier models parameterize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant.>

@@

Regardless of whether these hypotheses are true, our results further support the results of females having greater variation in the number of crossovers across cells in mice and humans [@lynn2002, @gruhn2013]. This feature of heterochiasmy supports evolutionary theories predicting distinct reproductive strategies between females (diversifying offspring) and males (maintain successful haplotypes) [@trivers2002].

@@

As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a highly conserved trait. The two locus modifier and SACE models predict sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks under linkage in males respectively.

The COM model, predicts the sex differences in the recombination landscape via interference is due to a combination of the length of the axis and differing strengths of chromosome movements during the pairing process [@hulten2011_COM] <@rubin2020mixing>.

@@

Our results from musculus^MSM^ and musculus^PWD^ show that the SC length and chromatin compaction is uncoupled from the direction of heterochiasmy. These results slightly depart from predictions which nominate chromatin organization (axis length) as the primary driver of heterochiasmy [@petkov2007]. While the relationship may hold for more general heterochiasmy patterns across broad scale comparisons, our results indicate the sexual dimorphism in chromatin compaction is conserved in house mouse (e.g. females have longer SC than males) and that the rapid male-specific evolution in crossover number per cell proceeded through another aspect of crossover regulation.

@@

Examples of evolution of interference strength are rare and may be complicated by changes in the underlying karyotype structure (Segura et al 2014). The general pattern of empirical measures of interference strength is that it has a negative correlation to the genome wide recombination rate [@ottoPaysuer2019]. This fits the logic of the chromosome axis acting as the 'real estate' for DSBs and aligns with known molecular mechanisms (cite). Two examples of the opposite direction, a positive correlation of interference strength and genome-wide recombination rate, involve descriptions of observations at the single bivalent level; the between lab-raised and wild mice of *Peromyscus leucopus* from [@peterson2019] or large effects on genome wide recombination rates across a short evolutionary differences; the within-sex comparison of two breeds of cattle with different genome wide recombination rates [@ma2015_cattle]. We propose that there is a difference in resolution and power between these two groups of empirical results with the positive and negative relationship between genome wide recombination rates and interference strength.

We propose that the **spindle based selection model** can explain the positive correlation via selection on the amount of sister cohesion connecting homologs. The evolution of genome wide recombination rates in our study is driven by a transition of the majority of 
bivalent having single crossovers to an enrichment of bivalents with two crossovers. All else being equal, interference strength would be expected to be equal or even weaker. Our results show that the two crossovers are spaced further apart when SC length is controlled for and chromosome size effects are minimized, in the strain enriched for two crossover bivalents. Assuming that chromatin compaction between pachtyene and metaphase is uniform along chromosomes, an outcome of this further spacing of two foci results in an increased area of sister cohesion connecting homologs (Figure X). This model gives rise to the possibility that the chromosome structure at metaphase is the target of selection  which is achieved by changing the average recombination landscape across bivalents within the cell.

The **COM** model predicts that interference and the recombination landscape arises from known oscillatory movements during prophase, however it lacks evolutionary based predictions. While the **SACE** modifier model does not explicitly model evolution of interference strength we note that a logical outcome of the main prediction of maintaining larger chromosome blocks in males, would be a recombination landscape with stronger interference strength.

We have focused on the model involving direct selection on the recombination pathway, which has an impact on the fitness of an individual, but we acknowledge connections to other models involving indirect selection on recombination rates which involve the fitness outcomes of offspring. A model from Goldstein et al, reviewed in [@ottoPaysuer2019] and @veller2019shuffle, predict that interference strength evolves whenever increased recombination rates evolve. The simulations under this model indicate that the space across multiple loci or between multiple crossovers increases in a positive manner with the genome wide recombination rate. A distinguishing feature of this model from other modifer models is that the number of crossovers per chromosome is kept constant. 
Empiracally, the range of crossovers per chromosome is quite limited, 1-3, for the majority of chromosomes [@ottoPaysuer2019, @stapley_variation_2017] across most taxa. Constraining the number of crossovers per chromosome in models may fit empirical data better than those where recombination rate across an abstracted genetic space is unconstrained.
